---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# fastpos

<!-- badges: start -->
[![Travis build status](https://travis-ci.org/johannes-titz/fastpos.svg?branch=master)](https://travis-ci.org/johannes-titz/fastpos)
[![Codecov test coverage](https://codecov.io/gh/johannes-titz/fastpos/branch/master/graph/badge.svg)](https://codecov.io/gh/johannes-titz/fastpos?branch=master)
<!-- badges: end -->

This package provides a fast algorithm to calculate the required sample size for a Pearson correlation to stabilize in the sequential framework of Schönbrodt and Perugini (2013; 2018). I assume that you have read the original paper or at least have an idea of how it works in principle. Essentially you want to find the sample size at which you can be sure that $1-\alpha$ percent of many studies would fall into a specified corridor of stability around an assumed population correlation and stay inside the corridor if you add more participants to your study. For instance, how many participants per study are required so that out of 100k studies, 90\% would fall into the region between .4 to .6 (a Pearson correlation) and not leave this region anymore when you add more participants (under the assumption that the population correlation is .5). This is also referred to as the critical point of stability.

## Installation

<!-- You can install the released version of fastpos from [CRAN](https://CRAN.R-project.org) with: -->

<!-- ``` r -->
<!-- install.packages("fastpos") -->
<!-- ``` -->
You can install the development version from [GitHub](https://github.com/) with devtools (and vignettes build, this takes a couple of seconds):

```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE, eval=FALSE}
devtools::install_github("johannes-titz/passt", build_vignettes = TRUE)
```

## Using *fastpos*
If you have found this page, I assume you either want (1) to calculate the critical point of stability for your own study or (2) to explore the method in general. If this is the case, read on and hopefully you will find what you are looking for. Let us first load the package and set a seed for reproducibility:

```{r setup}
library(fastpos)
set.seed(19950521)
```

In most cases you will just need the function **find_critical_pos** which will you give you the critical point of stability for your specific parameters.

Let us reproduce Schönbrodt and Peruigini's quite famous, often cited table of the critical points of stability for a precision of 0.1. We reduce the number of studies to 10k so that it runs fairly quickly.

```{r message=TRUE, warning=TRUE, paged.print=TRUE}
find_critical_pos(rho = seq(.1, .7, .1), sample_size_max = 1000,
                  n_studies = 10000)
```

Note that a warning is shown because in some simulations the corridor of stability was not reached. As long as this number is low, this should not affect the estimates much. But if you want to get more accurate estimates increase the maximum sample size.

You can obviously parallelize the process, which will be especially useful if you want to run many simulations. For instance, if you increase the number of studies to 100k (as in the original article), it will take less than a minute on a modern CPU with several cores. On my i7-2640 with 4 cores, it takes about 30 s. Overall, this is a speedup of more than 1000 compared to Schönbrodt and Perugini's code.

If you want to dig deeper, you can have a look at the functions that *find_critical_pos* builds upon. **simulate_pos** is the workhorse of the package. It calls a C++ function to calculate correlations sequentially and it does this pretty fast (but you know that already, right?). A rawish approach would be to create a population with **create_pop** and pass it to **simulate_pos**:

```{r, fig.width = 6, fig.height = 4.2}
pop <- create_pop(0.5, 1000000)
pos <- simulate_pos(x_pop = pop[,1],
                    y_pop = pop[,2],
                    number_of_studies = 1000,
                    sample_size_min = 20,
                    sample_size_max = 1000,
                    replace = T,
                    lower_limit = 0.4,
                    upper_limit = 0.6)
hist(pos, xlim = c(0, 1000), xlab = c("Point of stability"),
     main = "Histogram of points of stability for rho = .5+-.1")
quantile(pos, c(.8, .9, .95))
```

Note that no warning message appears if the corridor is not reached. It will simply return the maximum sample size. So pay careful attention if you work with this function and adjust the maximum sample size as needed.

**create_pop** creates the population matrix by using **mvrnorm**. This is a much simpler way compared to Schönbrodt and Perugini's approach, but the results do not seem to differ. If you are interested in how population parameters (e.g. skewness) affect the point of stability, you should rather refer to the population generating functions in Schönbrodt and Perugini's work.

## Issues and Support
If you find any bugs, please use the issue tracker at:

https://github.com/johannes-titz/fastpos/issues

If you need assistance in how to use the package, drop me an e-mail at johannes at titz.science or johannes.titz at gmail.com

## Contributing
Contributions of any kind are very welcome! I will sincerely consider every suggestion on how to improve the code, the documentation and the presented examples. Even minor things, such as suggestions for better wording or improving grammar in any part of the package are considered as valuable contributions.

If you want to make a pull request, please check that you can still build the package without any errors, warnings or notes. Overall, simply stick to the R packages book: https://r-pkgs.org/ and follow the code style described here: http://r-pkgs.had.co.nz/r.html#style

## References
Schönbrodt, F. D. & Perugini, M. (2013). At what sample size do correlations stabilize? *Journal of Research in Personality, 47*, 609-612. [https://doi.org/10.1016/j.jrp.2013.05.009]
 
Schönbrodt, F. D. & Perugini, M. (2018) Corrigendum to “At what sample size do correlations stabilize?” [J. Res. Pers. 47 (2013) 609–612. https://doi.org/10.1016/j.jrp.2013.05.009]. *Journal of Research in Personality, 74*, 194. [https://doi.org/10.1016/j.jrp.2018.02.010]

